Demography and linked selection in wild and domesticated maize

### 2015-08-28 manuscript.tex
minor edits

Our initial comparison of synonymous and missense substitutions has low power to detect the effects of selection acting on multiple mutations or standing genetic variation, because in such cases diversity is not necessarily reduced \cite{innan2004,messer2013}. Such ``soft sweeps'', however, are still expected to occur more frequently in functional regions of the genome and could provide an alternative explanation for the observed reduction of diversity in genes. To test this possibility, we performed a genome-wide scan for selection using a method expected to be reasonably sensitive to both hard and soft sweeps \cite{garud2015}; 
after removing genes in the top 20\% of the H12 statistic used to identify targets of selection, we observe that qualitative differences between maize and teosinte in patterns of diversity in and near genes remained unchanged (Figure \ref{sFig:H12}A).

We interpret these combined results as suggesting that purifying selection has left a more pronounced signature in the teosinte genome due to the increased efficacy of selection resulting from differences in effective population size.

@@

\subsection{Demography of maize domestication}
To explore whether differences in the efficacy of purifying selection between maize and teosinte can be explained by demographic processes, we estimated the parameters of a simple domestication bottleneck model  (Figure \ref{fig:bottleneck}). 
The most likely model estimates an ancestral population mutation rate of $\theta=0.0147$ per bp, which translates to an effective population size of $N_a \approx 123,000$ individuals given the mutation rate \cite{clark2005}.
The maize population splits from teosinte $\approx 15,000$ generations in the past with an initial size of only $\approx 5\% $ of ancestral $N_a$. 
Following domestication we estimate considerable gene flow between the populations: $M_{tm} = N_a \times 1.1 \times 10^{-5}$  migrants per generation from teosinte to maize and $M_{mt} = N_a \times 1.4 \times 10^{-5}$ migrants from maize to teosinte. 
After its split from teosinte our model posits exponential population growth in maize, estimating a final modern effective population size of $N_m \approx 370,000$.

@@

\color{blue}
Finally, we reconstructed the demographic history of maize and teosinte using a recently developed coalescent approach \citex that does not require conditioning on a particular model (Figure \X).
Though this analysis does highlight a potential weakness with our assumption of a constant size for teosinte, it nonetheless is broadly consistent with our simple model,   identifying a clear domestication bottleneck followed by rapid population expansion (Figure \X). 
\color{black}

\subsection{Population expansion leads to stronger purifying selection in modern maize}
Motivated by the rapid post-domestication expansion of maize evident from our demographic analysis, we investigated whether patterns of diversity at low frequency --- and therefore on average younger --- polymorphisms would show similar results in comparisons of maize and teosinte.
We thus repeated our analysis of diversity using alleles present in only a single individual in the sample.
In direct contrast to patterns observed for pairwise nucleotide diversity (Figure \ref{fig:purify}A), diversity of singletons was more strongly reduced in and near genes in maize than in teosinte (Figure \ref{fig:purify}B), even after downsampling our maize data to account for differences in sample size. This relationship was maintained when we analyzed diversity surrounding genes that did not show evidence of hard or soft sweeps according to H12 (Figure \ref{sFig:H12})
Finally, while direct comparison of curves for pairwise and singleton diversity within taxa are consistent with at least some non-equilibrium dynamics in teosinte, as indicated by our coalescent-based demographic analysis, these too reveal much stronger differences in maize (Figure \ref{sFig:singletonPi}).

@@

Although we estimate that the modern effective size of maize is three times that of teosinte, the small size of our sample dramatically reduces our power to estimate rapid population growth since rapid growth is most evidenced by rare alleles\cite{keinan2012}.  
A better estimate of the effective size of modern maize comes from our analysis of singleton diversity in a large sample of maize landraces \cite{Hearne2015}.
These data suggest a modern effective size of nearly 1 million, nearly eight times higher than modern teosinte.
Nonetheless, missing data, sample size, and ascertainment bias in the genotyping data all bias this estimate downward as well.
We can, however, get reasonable estimates of the census size of modern maize.
There are 47.9 million ha of open-pollinated maize in production \cite{cimmyt1999}, with a lower limit of 25,000 individuals planter per hectare \cite{baden2001culture}.
Using this lower bound on planting density, ignoring all industrial maize agriculture, and even assuming the effective size is only 0.1\% of the census size, this still implies a modern effective population size of more than one billion.
Clearly, the effective size of modern maize is extremely large.

\subsection{Demography influences the efficiency of purifying selection}
The observation that maize pairwise diversity is less impacted by the distance to the nearest gene than is teosinte pairwise diversity is reasonable from a long-term evolutionary standpoint. Purifying selection works more efficiently in a large population than in a small one \cite{kimura1984}, so this observation likely reflects that prediction. Since maize bottlenecked before recovering exponentially, the average $N_e$ of maize over the previous several thousand generations is much smaller than that of teosinte. Therefore, our observation shows that purifying selection in maize has not purged deleterious alleles, or the neutral alleles they are linked to, as effectively as in teosinte.

@@

\color{blue}
Finally, MSMC text here
\color{black}

@@

\begin{acknowledgments}
We are indebted to Graham Coop and Simon Aeshbacher for their constructive input during this study. \textcolor{red}{We thank Robert Bukowski and Qi Sun for providing early-access Panzea data.} Funding was provided by NSF Plant Genome Research Project 1238014.
\end{acknowledgments}

>>>>

Our initial comparison of synonymous and missense substitutions has low power to detect the effects of selection acting on multiple mutations or standing genetic variation, because in such cases diversity is not necessarily reduced \cite{innan2004,messer2013}. 
Such ``soft sweeps'', however, are still expected to occur more frequently in functional regions of the genome and could provide an alternative explanation for the observed reduction of diversity in genes. 
To test this possibility, we performed a genome-wide scan for selection using a method expected to be reasonably sensitive to both hard and soft sweeps \cite{garud2015}. 
After removing genes in the top 20\% of the H12 statistic used to identify targets of selection, we observe that qualitative differences between maize and teosinte in patterns of diversity in and near genes remained unchanged (Figure \ref{sFig:H12}A).

We interpret these combined results as suggesting that purifying selection has left a more pronounced signature in the teosinte genome due to the increased efficacy of selection resulting from differences in effective population size.\jri{weird that this is a 1-sentence paragraph?}

@@

\subsection{Demography of maize domestication}
To explore whether differences in the efficacy of purifying selection between maize and teosinte can be explained by demographic processes, we estimated the parameters of a simple domestication bottleneck model  (Figure \ref{fig:bottleneck}). 
The most likely model estimates an ancestral population mutation rate of $\theta=0.0147$ per bp, which translates to an effective population size of $N_a \approx 123,000$ individuals given the mutation rate \cite{clark2005}.
The maize population splits from teosinte $\approx 15,000$ generations in the past with an initial size of only $\approx 5\% $ of ancestral $N_a$. 
Following domestication we estimate considerable gene flow between the populations: $M_{tm} =  1.1 \times 10^{-5} \times N_a $  migrants per generation from teosinte to maize and $M_{mt} =  1.4 \times 10^{-5} \times N_a$ migrants from maize to teosinte. 
After its split from teosinte our model posits exponential population growth in maize, estimating a final modern effective population size of $N_m \approx 370,000$.

@@

%MSMC results
%Finally, we reconstructed the demographic history of maize and teosinte using a recently developed coalescent approach \citex that does not require conditioning on a particular model (Figure \X).
%Though this analysis does highlight a potential weakness with our assumption of a constant size for teosinte, it nonetheless is broadly consistent with our simple model,   identifying a clear domestication bottleneck followed by rapid population expansion (Figure \X).

\subsection{Population expansion leads to stronger purifying selection in modern maize}
Motivated by the rapid post-domestication expansion of maize evident from our demographic analysis, we investigated whether patterns of diversity at low frequency --- and therefore on average younger --- polymorphisms would show similar results in comparisons of maize and teosinte.
We thus repeated our analysis of diversity using alleles present in only a single individual in the sample.
In direct contrast to patterns observed for pairwise nucleotide diversity (Figure \ref{fig:purify}A), diversity of singletons was more strongly reduced in and near genes in maize than in teosinte (Figure \ref{fig:purify}B), even after downsampling our maize data to account for differences in sample size. This relationship was maintained when we analyzed diversity surrounding genes that did not show evidence of hard or soft sweeps according to H12 (Figure \ref{sFig:H12})
Finally, while direct comparison of curves for pairwise and singleton diversity within taxa are consistent with at least some non-equilibrium dynamics in teosinte, these too reveal much stronger differences in maize (Figure \ref{sFig:singletonPi}). %ref. here to MSMC if we use: "as indicated by our coalescent-based demographic analysis"

@@

Although we estimate that the modern effective size of maize is three times that of teosinte, the small size of our sample dramatically reduces our power to identify low frequency alleles most sensitive to rapid population growth \cite{keinan2012}.  
A better estimate of the effective size of modern maize comes from our analysis of singleton diversity in a large sample of maize landraces \cite{Hearne2015}.
These data suggest a modern effective size of nearly 1 million, nearly eight times higher than modern teosinte.
Nonetheless, missing data, sample size, and ascertainment bias in the genotyping data all bias this estimate downward as well.
We can, however, get reasonable estimates of the census size of modern maize.
There are 47.9 million ha of open-pollinated maize in production \cite{cimmyt1999}, with a lower limit of 25,000 individuals planter per hectare \cite{baden2001culture}.
Using this lower bound on planting density, ignoring all industrial maize agriculture, and even assuming the effective size is only 0.1\% of the census size, this still implies a modern effective population size of more than one billion.
Clearly, the effective size of modern maize is extremely large.

\subsection{Demography influences the efficiency of purifying selection}
The observation that pairwise diversity in maize is less affected by the distance to the nearest gene than in teosinte is reasonable from a long-term evolutionary standpoint. Purifying selection works more efficiently in a large population than in a small one \cite{kimura1984}, so this observation likely reflects that prediction. Since maize bottlenecked before recovering exponentially, the average $N_e$ of maize over the previous several thousand generations is much smaller than that of teosinte. Therefore, our observation shows that purifying selection in maize has not purged deleterious alleles, or the neutral alleles they are linked to, as effectively as in teosinte.

@@

%%%Shorten if we include MSMC
%Finally, MSMC \cite{schiffels2014} was employed to complement our model-based demographic inference. 
%Using six each of maize and teosinte
%The analyses were run for the following two groups: six maize haplotypes (BKN022, BKN025, BKN029, BKN030, BKN031, BKN033) and six teosinte haplotypes (TIL01, TIL03, TIL09, TIL10, TIL11 and TIL14).
%SNPs were called in ANGSD \cite{korneliussen2014} with the SNP p value set as $1e-6$.
%As the samples are highly inbred, each individual was treated as one haplotype.
%Only Sites with mapping quality bigger than $30$, base quality over $20$ and depth between half and twice of its mean depth were chosen for the analyses, and the residual heterozygous sites were removed from the analyses.
%In addition, as the maize genome contains a large body of repetitive regions \cite{schnable2009}, we used Heng Li's software SNPable (\href{<url>}{<http://lh3lh3.users.sourceforge.net/snpable.shtml>}) to mask out genomic regions, on which short sequencing reads are not uniquely mapped.
%The mappability mask file was generated by cutting the maize genome with $1bp$ shift each time into the $100bp$ single-end reads, which were then mapped back to the maize B73 reference genome \cite{schnable2009}.
%The sites with the majority of overlapping $100$-mers mapped uniquely without one mismatch were called the "SNPable" sites and used for the MSMC analyses.
%The pattern parameter $20\times2+20\times4+10\times2$ was used for the population size inference.

@@

\begin{acknowledgments}
We are indebted to Graham Coop and Simon Aeshbacher for their constructive input during this study. We thank Robert Bukowski and Qi Sun for providing early-access data from maize HapMap3. Funding was provided by NSF Plant Genome Research Project 1238014.
\end{acknowledgments}
